One exact and one random sampling method to assess the reliability of inferences from network enrichment analysis

### 2018-05-08 doc/genesetdp.tex
x

In network-based gene set analysis one score a query set of genes, $ \mathcal{Q}=\{g_1 \ldots g_Q\}$, from a genome with genes, $\mathcal{G}=\{g_1 \ldots g_G\}$ ($Q \le G$), and how they relate to a pathway $\{p_1 \ldots p_P\}$. In NEA the pathway maps to the genome through a network ${x_{ij}}$, where $x_{ij}=1$ if $g_i$ and $p_j$ are connected, or $x_{ij}=0$ otherwise. The pathways are scored by summing up all connections, i.e. the pathway score is

@@

\begin{equation}
p(s)=\frac{N(s)/2 +\sum_{s'=s+1}^{S} N(s')}{\sum_{s'=0}^{S} N(s')},
\label{eq:pval}
\end{equation}
where $S$ is the maximal score $s$ that $Q$ genes can obtain from Equation \ref{eq:sum_ij}.

@@

We first implemented a random sampling algorithm to assess $N(s)$. We randomly selected sets of $Q$ genes from $\mathcal{G}$. These sets scores were calculated with Equation \ref{eq:sum_i}. We counted the number of times $F_B(s)$ a score of $s$ was obtained when sampling $B$ gene sets. We see that $F_B(s)$ will approximately have the same shape as $N(s)$ when selecting a large $B$, and we can hence calculate $p$~values by replacing $N$ with $F_B$ in Equation~\ref{eq:pval}. We refer to this procedure as GeneSetMC.

@@

Source code of an implementation of GeneSetDP and GeneSetMC, as well as code for generating the plots of this paper, is available from \url{https://github.com/statisticalbiotechnology/genesetdp}.

>>>>

In network-based gene set analysis one score a query set of genes, $ \mathcal{Q}=\{g_1 \ldots g_Q\}$, from a genome with genes, $\mathcal{G}=\{g_1 \ldots g_G\}$ ($Q \le G$), and how they relate to a pathway $\{p_1 \ldots p_P\}$. In NEA the pathway maps to the genome through a network, ${x_{ij}}$, where $x_{ij}=1$ if $g_i$ and $p_j$ are connected, or $x_{ij}=0$ otherwise. The pathways are scored by summing up all connections, i.e. the pathway score can be expressed as,

@@

\begin{equation}
p(s)=\frac{N(s)/2 +\sum_{s'=s+1}^{S} N(s')}{\sum_{s'=0}^{S} N(s')},
\label{eq:pval}
\end{equation}
where $S$ is the maximal score, $s$, that $Q$ genes can obtain from Equation \ref{eq:sum_ij}.

@@

We first implemented a random sampling algorithm to assess $N(s)$. We randomly selected sets of $Q$ genes from $\mathcal{G}$. These sets scores were calculated with Equation \ref{eq:sum_i}. We counted the number of times, $F_B(s)$, a score of $s$ was obtained when sampling $B$ gene sets. We see that $F_B(s)$ will approximately have the same shape as $N(s)$ when selecting a large $B$, and we can hence calculate $p$~values by replacing $N$ with $F_B$ in Equation~\ref{eq:pval}. We refer to this procedure as GeneSetMC.

@@

The source code of an implementation of GeneSetDP and GeneSetMC, as well as code for generating the plots of this paper, is available from \url{https://github.com/statisticalbiotechnology/genesetdp}.
